Liveness-Based Garbage Collection for Lazy Languages

### 2015-11-30 paper/paper.tex
fixing reviewr 1 comments

The  variable \pz\   returned  at  $\psi_3$ depends  on  \py\ 
  through  the  expression $(+~\py~1)$.   $\py$  in  turn depends  on  the
  closure  variable  $\px_{\mathit{\pi}_1}$   through  $(\length~\px_{\mathit{\pi}_1})$.   We
  denote  this chain  of dependences  as $\lbrack  \psi_3\!\!\!:\!\!\!\pz \leftarrow
  (+~\py~1),\,  \py  \leftarrow   (\length~\px_{\mathit{\pi}_1})\rbrack$.   Indeed,  the
  closures  in  the  heap  are   a  runtime  representation  of  these
  dependences.  Since  $\pz$  is  evaluated   at  $\psi_3$  due  to  the
  expression $\SRETURN~\pz$,  the demand made by  the calling context(s)
  of  $f$ places  a demand  on  $\pz$ which  will impart  a liveness  to
  $\px_{\pi_1}$.  Other dependence chains which result in a liveness for
  $\px_{\pi_1}$       are       $\lbrack       \psi_1\!\!\!:\!\!\py       \leftarrow
  (\length~\px_{\pi_1})\rbrack$ and $\lbrack \psi_2\!\!\!:\!\!\!\pw \leftarrow (/~\pu~\py),\,\,
  \py  \leftarrow  (\length~\px_{\pi_1})\rbrack$.   The  liveness  analysis
  described in this section declares the liveness of $\px_{\pi_1}$ to be
  a union of the liveness arising  out of these dependence chains.  To
  be safe, a  garbage collection during evaluation of  $\py$ at $\psi_1$
  has  to  use  this  liveness  to copy  the  heap  from  $\px_{\pi_1}$.
  However,  notice that  if  a garbage  collection  takes place  while
  evaluating $\pz$ at $\psi_3$, it can safely consider only the liveness
  arising  out of  the dependence  chain\linebreak $\lbrack  \psi_3\!\!\!:\!\!\!\pz \leftarrow
  (+~\py~1),\,   \py  \leftarrow   (\length~\x_{\pi_1})\rbrack$.   The   garbage
  collection  scheme described  in Section~\ref{sec:GC-scheme}  uses a
  generalization  of   this  observation   to  dynamically   select  a
  evaluation point specific liveness in order to collect more garbage.

@@

In  summary, the  equations  generated during  liveness
analysis are:
\begin{enumerate}
\item   For  each   function  $\mathit{f}$,   equations
  defining \Df{f}{i} for use by \Lfonly$_{\mathit f}$.
\item For each function $\mathit{f}$, an equation defining the summary
  demand $\sigma_{\mathit f}$ on $e_f$.
\item   For  each   function  $\mathit{f}$   (including
  $\mainpgm$) an  equation  defining
  liveness for each garbage in  $e_{\mathit f}$.
\end{enumerate}
\subsection{Solving liveness equations---the grammar
interpretation}\label{sec:grammar-formulation}      The
equations  above   can  now  be  re-interpreted   as  a
context-free    grammar   (CFG)    on   the    alphabet
$\lbrace\acar,  \acdr,  \bcar,  \bcdr,  \clazy\rbrace$.
Let \var{$X$}  denote the  non-terminal for  a variable
$X$  occurring on  the LHS  of the  equations generated
from  the  analysis.  We  can  think  of the  resulting
productions as being  associated with several grammars,
one for each non-terminal \var{\Lanv{x}{}} regarded as
a  start  symbol.   As  an  example,  the  grammar  for
\var{\Lanv{\xl}{}} and \var{\Lanv{\pa}{}}    comprise     the     following
productions:
\begin{eqnarray*}
  \var{\Lanv{\xl}{}}  &\rightarrow& 
  \clazy\var{\sigma_{\length}} \mid \acdr \var{\Df {\length}{1}}{\clazy\var{\sigma_{\length}}}  \\
  & & \mid
  \clazy\var{\Df{\length}{1}}{\clazy\var{\sigma_{\length}}} \\
  \var{\Df{\length}{1}} &\rightarrow& \clazy \mid
  \acdr\var{\Df{\length}{1}}\clazy
       \mid \clazy\var{\Df{\length}{1}}{\clazy}\\
\langle {\sigma_{\length}} \rangle
&\rightarrow&
\epsilon  \mid \clazy\langle{\sigma_{\length}}\rangle \\
\var{\Lanv{\pa}{}} &\rightarrow& \clazy \bcar \var{\Df{\length}{1}}
\end{eqnarray*}
Other  equations  can   be  converted  similarly.   The
language   generated   by  \var{\Lanv{x}{}},   denoted
$\mathscr{L}(\var{\Lanv{x}{}})$,   is    the   desired
solution  of  \Lanv{x}{}.    
However, recall that the decision problem that we are interested in
during garbage collection is the following: 
\begin{quote}
Let $x.\alpha$ be a forward access path consisting only
of edges  \acar\ and \acdr\  (but not \bcar,  \bcdr\ or
\clazy).       Let      $\mathscr{L}(\var{\Lanv{x}{}})
\stackrel{*}{\hookrightarrow}  \sigma$, where  $\sigma$
consists  of  forward  access  paths  only.  Then  does
$\alpha \in \sigma$?
\end{quote}

@@

The strongly regular grammar is converted  into a set of NFAs, one for
each $\var{\Lanv{x}{}}$.  The  $\hookrightarrow$ simplification is now
done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
pairs  of consecutive  edges  labeled \bcar\acar\  or \bcdr\acdr\  and
constructing  the $\epsilon$-closure  till a  fixed point  is reached,
after which  the edges labeled  \bcar \  and \bcdr\ are  deleted.  The
details of the  algorithm, its correctness and  termination proofs are
given  in~\cite{karkare07liveness,asati14lgc}.   The
resulting  automaton   has  edges  labeled  with   \acar,  \acdr\  and
\clazy\ only.  In this  automaton, for every  edge labeled  \clazy, we
check if the source node of the edge has a path to a final 
state.  If yes, \warning{we mark the mark the source node as final. Finally, we
remove all the edges labeled \clazy\  and convert the automaton into a
deterministic    automaton.}    These steps   effectively   implement    the
$\hookrightarrow$   simplification  rules   for   \bcar,  \bcdr,   and
\clazy\ to  obtain forward paths.  While checking for  liveness during
garbage  collection, a  forward access-path  is valid  only if  it can
reach a  final state.  Figure~\ref{fig:example-automata}(a)  shows the
NFA that is obtained from  the grammar for \var{\Df{\length}{1}}, and   the
final  DFA  is  shown in  Figure~\ref{fig:example-automata}(b).   This
expectedly  says that for a  demand  \warning{$\epsilon$  on  \length},  the
liveness of  its argument  is $\acdr^{*}$  (the spine  of the  list is
traversed).  Similarly, Figure~\ref{fig:example-automata}(c) shows the
NFA     for     \var{\Lanv{\pa}{}}.      The     DFA     shown     in
Figure~\ref{fig:example-automata}(d)  does  not   accept  any  forward
paths.   This  reflects  the  lazy  nature  of  our  language.   Since
\length\ does not  evaluate the member elements of  the argument list,
the  closure for  \pa\ is  never evaluated  and is  reclaimed whenever
liveness based garbage collection is triggered beyond $\pi_9$.

>>>>

The  variable \pz\   returned  at  $\psi_3$ depends  on  \py\ 
  through  the  expression $(+~\py~1)$.   $\py$  in  turn depends  on  the
  closure  variable  $\px_{\mathit{\pi}_1}$   through  $(\length~\px_{\mathit{\pi}_1})$.   We
  denote  this chain  of dependences  as $\lbrack  \psi_3\!\!\!:\!\!\!\pz \leftarrow
  (+~\py~1),\,  \py  \leftarrow   (\length~\px_{\mathit{\pi}_1})\rbrack$.   Indeed,  the
  closures  in  the  heap  are   a  runtime  representation  of  these
  dependences.  Since  $\pz$  is  evaluated   at  $\psi_3$  due  to  the
  expression $\SRETURN~\pz$,  the demand made by  the calling context(s)
  of  $f$ places  a demand  on  $\pz$ which  will impart  a liveness  to
  $\px_{\pi_1}$.  Other dependence chains which result in a liveness for
  $\px_{\pi_1}$       are       $\lbrack       \psi_1\!\!\!:\!\!\py       \leftarrow
  (\length~\px_{\pi_1})\rbrack$ and $\lbrack \psi_2\!\!\!:\!\!\!\pw \leftarrow (/~\pu~\py),\,\,
  \py  \leftarrow  (\length~\px_{\pi_1})\rbrack$.   The  liveness  analysis
  described in this section declares the liveness of $\px_{\pi_1}$ to be
  a union of the liveness arising from these dependence chains.  To
  be safe, a  garbage collection during evaluation of  $\py$ at $\psi_1$
  has  to  use  this  liveness  to copy  the  heap  from  $\px_{\pi_1}$.
  However,  notice that  if  a garbage  collection  takes place  while
  evaluating $\pz$ at $\psi_3$, it can safely consider only the liveness
  arising from the dependence  chain\linebreak $\lbrack  \psi_3\!\!\!:\!\!\!\pz \leftarrow
  (+~\py~1),\,   \py  \leftarrow   (\length~\x_{\pi_1})\rbrack$.   The   garbage
  collection  scheme described  in Section~\ref{sec:GC-scheme}  uses a
  generalization  of   this  observation   to  dynamically   select  a
  evaluation point specific liveness in order to collect more garbage.

@@

In  summary, the  equations  generated during  liveness
analysis are:
\begin{enumerate}
\item   For  each   function  $\mathit{f}$,   equations
  defining \Df{f}{i} for use by \Lfonly$_{\mathit f}$.
\item For each function $\mathit{f}$, an equation defining the summary
  demand $\sigma_{\mathit f}$ on $e_f$.
\item   For  each   function  $\mathit{f}$   (including
  $\mainpgm$) an  equation  defining
  liveness for each garbage in  $e_{\mathit f}$.
\end{enumerate}
\subsection{Solving liveness equations---the grammar
interpretation}\label{sec:grammar-formulation}      The
equations  above   can  now  be  re-interpreted   as  a
context-free    grammar   (CFG)    on   the    alphabet
$\lbrace\acar,  \acdr,  \bcar,  \bcdr,  \clazy\rbrace$.
Let \var{$X$}  denote the  non-terminal for  a variable
$X$  occurring on  the LHS  of the  equations generated
from  the  analysis.  We  can  think  of the  resulting
productions as being  associated with several grammars,
one for each non-terminal \var{\Lanv{x}{}} regarded as
a  start  symbol.   As  an  example,  the  grammar  for
\var{\Lanv{\xl}{}} and \var{\Lanv{\pa}{}}    comprise     the     following
productions:
\begin{eqnarray*}
  \var{\Lanv{\xl}{}}  &\rightarrow& 
  \clazy\var{\sigma_{\length}} \mid \acdr \var{\Df {\length}{1}}{\clazy\var{\sigma_{\length}}}  \\
  & & \mid
  \clazy\var{\Df{\length}{1}}{\clazy\var{\sigma_{\length}}} \\
  \var{\Df{\length}{1}} &\rightarrow& \clazy \mid
  \acdr\var{\Df{\length}{1}}\clazy
       \mid \clazy\var{\Df{\length}{1}}{\clazy}\\
\langle {\sigma_{\length}} \rangle
&\rightarrow&
\epsilon  \mid \clazy\langle{\sigma_{\length}}\rangle \\
\var{\Lanv{\pa}{}} &\rightarrow& \clazy \bcar \var{\Df{\length}{1}}
\end{eqnarray*}
Other  equations  can   be  converted  similarly.   The
language   generated   by  \var{\Lanv{x}{}},   denoted
$\mathscr{L}(\var{\Lanv{x}{}})$,   is    the   desired
solution  of  \Lanv{x}{}.    
However, recall that the decision problem that we are interested in
during garbage collection is the following: 
\begin{quote}
Let $x.\alpha$ be a forward access path (consisting only
of edges  \acar\ and \acdr\  but not \bcar,  \bcdr\ or
\clazy).       Let      $\mathscr{L}(\var{\Lanv{x}{}})
\stackrel{*}{\hookrightarrow}  \sigma$, where  $\sigma$
consists  of  forward  access  paths  only.  Then  does
$\alpha \in \sigma$?
\end{quote}

@@

The strongly regular grammar is converted  into a set of NFAs, one for
each $\var{\Lanv{x}{}}$.  The  $\hookrightarrow$ simplification is now
done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
pairs  of consecutive  edges  labeled \bcar\acar\  or \bcdr\acdr\  and
constructing  the $\epsilon$-closure  till a  fixed point  is reached,
after which  the edges labeled  \bcar \  and \bcdr\ are  deleted.  The
details of the  algorithm, its correctness and  termination proofs are
given  in~\cite{karkare07liveness,asati14lgc}.   The
resulting  automaton   has  edges  labeled  with   \acar,  \acdr\  and
\clazy\ only.  In this  automaton, for every  edge labeled  \clazy, we
check if the source node of the edge has a path to a final 
state.  If yes, \warning{we mark the mark the source node as final. Finally, we
remove all the edges labeled \clazy\  and convert the automaton into a
deterministic    automaton.}    These steps   effectively   implement    the
$\hookrightarrow$   simplification  rules   for   \bcar,  \bcdr,   and
\clazy\ to  obtain forward access paths.  While checking for  liveness during
garbage  collection, a  forward access path  is valid  only if  it can
reach a  final state.  Figure~\ref{fig:example-automata}(a)  shows the
NFA that is obtained from  the grammar for \var{\Df{\length}{1}}, and   the
final  DFA  is  shown in  Figure~\ref{fig:example-automata}(b).   This
expectedly  says that for a  demand  \warning{$\epsilon$  on  \length},  the
liveness of  its argument  is $\acdr^{*}$  (the spine  of the  list is
traversed).  Similarly, Figure~\ref{fig:example-automata}(c) shows the
NFA     for     \var{\Lanv{\pa}{}}.      The     DFA     shown     in
Figure~\ref{fig:example-automata}(d)  does  not   accept  any  forward
paths.   This  reflects  the  lazy  nature  of  our  language.   Since
\length\ does not  evaluate the member elements of  the argument list,
the  closure for  \pa\ is  never evaluated  and is  reclaimed whenever
liveness based garbage collection is triggered beyond $\pi_9$.
